Comparing human and model-based forecasts of COVID-19 in Germany and Poland

### 2022-05-19 manuscript/manuscript.tex
changes for manuscript revision

Data on daily reported test positive cases and deaths linked to COVID-19 were provided by the organisers of the German and Polish Forecast hub. Until December 14th, 2020, these data were sourced from the European Centre for Disease Control (33). After ECDC stopped publishing daily data, observations were sourced from the Robert Koch Institute (RKI) and the Polish Ministry of Health for the remainder of the submission period (34). These data are subject to reporting artefacts, (such as for example delayed case reporting in Poland on the 24th November, (35)), changes in reporting over time, and variation in testing regimes (for example in Germany from the 11th of November on, (36)). The ECDC data as well as the data published by the Polish Ministry of Health were also subject to data revisions, although most of them (with a notable exception of a data update for October 12 2020 in Germany) only affected daily, not weekly data (see Figures \ref{fig:daily-truth-update} and \ref{fig:weekly-truth-update})

@@

Our crowd forecasts were created as an ensemble of forecasts made by individual participants every week through a web application (\url{https://cmmid-lshtm.shinyapps.io/crowd-forecast/}). Weekly forecasts had to be submitted before Tuesday 12pm every week, but participants were asked to only use any information or data that was already available by Monday night. The application was built using the \texttt{shiny} and \texttt{golem} R packages (37,38) and is available in the \texttt{crowdforecastr} R package (31). To make a forecast in the application participants could select a predictive distribution (with the default being log-normal) to represent the probability that the forecasted quantity took certain values. Median and width of the uncertainty could be adjusted by either interacting with a figure showing their forecast or providing numerical values (see screenshot in Figure \ref{fig:screenshot} in the SI). The default shown was a repetition of the last known observation with constant uncertainty around it computed as the standard deviation of the last four changes in weekly log observed forecasts (i.e.~as \(\sigma(log(value4) - log(value3), log(value3) - log(value2), \ldots )\)).
Our interface also allowed participants to view past observations based on the hub data, as well as their forecasts, on a logarithmic scale and presented additional contextual COVID-19 data sourced from (39). These data included, for example, notifications of both test positive COVID-19 cases and COVID-19 linked deaths and the number of COVID-19 tests conducted over time.

Forecasts were stored in a Google Sheet and downloaded, cleaned and processed every week for submission to the Forecast Hub. If a forecaster had submitted multiple predictions for a single target, only the latest submission was kept. Information on the chosen distribution as well as the parameters for median and width were used to obtain the required set of 23 quantiles from that distribution. Forecasts from all forecasters were then aggregated using an unweighted quantile-wise mean (i.e., the \(\alpha\)-quantile of the ensemble is given by the mean of all submitted \(\alpha\)-quantiles). On a few occasions, individual forecasts were assessed as clearly erroneous by visual inspection and subsequently removed before aggregation and were excluded from the submission as well as the analysis.

@@

All forecasting approaches, including the Hub ensemble, were overconfident, i.e.~they showed lower than nominal coverage (meaning that 50\% (90\%) prediction intervals generally covered less than 50\% (90\%) of the actually observed values) (Figure \ref{fig:agg-performance-all}C and \ref{fig:agg-performance-all}D). Coverage for all forecasts deteriorated with increasing forecast horizon, indicating that all forecasting approaches struggled to quantify uncertainty appropriately for case forecasts. This was especially an issue for crowd forecasts, which had markedly shorter prediction intervals (i.e., narrower and more confident predictive distributions) than other approaches (Figure \ref{fig:agg-performance-all}E) and only showed a small increase in uncertainty across forecast horizons. In spite of good performance in terms of the absolute error (Figure \ref{fig:agg-performance-all}G), the narrow forecast intervals led to forecasts which were severely overconfident (covering only 36\% and 55\% of all observations with the 50\% and 90\% prediction intervals of all forecasts made at a two week forecast horizon, and only 5\% and 38\% four weeks ahead) (Figure \ref{fig:agg-performance-all}C,D and Tables \ref{tab:score-table-2} and \ref{tab:score-table-4}). Despite worse performance in terms of absolute error (Figure \ref{fig:agg-performance-all}G), the renewal model achieved better calibration (comparable to the Hub ensemble), as uncertainty increased rapidly across forecast horizons.

@@

Generally, trends in overall performance were broadly similar across locations (Figures \ref{fig:performance-locations} and \ref{fig:performance-locations-rel}).
Due to the differing population sizes and numbers of notifications in Germany and Poland absolute scores were difficult to compare directly. However, relative to the Hub ensemble, the crowd forecasts performed noticeably better in Germany than in Poland and the renewal model better in Poland than in Germany (Figure \ref{fig:performance-locations-rel}A and \ref{fig:performance-locations-rel}G).

@@

\begin{figure}[H]
\includegraphics[width=1\linewidth,]{../analysis/plots/aggregate-performance-rel-ensemble-v4} \caption{Visualisation of relative aggregate performance metrics across forecast horizons for the different versions of the Hub median ensemble. “Hub-ensemble” *excludes* all our models, Hub-ensemble-all *includes* all of our models, “Hub-ensemble-real” is the actual hub-ensemble observed in reality, which includes the renewal model and the crowd forecasts, but not the convolution model. A, B: mean weighted interval score (WIS) across horizons relative to the Hub ensemble (lower values indicate better performance). C, D: Empirical coverage of the 50\% and 90\% prediction intervals minus empirical coverage observed for the Hub ensemble. E: Dispersion relative to the dispersion of the Hub ensemble. Higher values mean greater dispersion of the forecast and imply ceteris paribus a worse score. F: Bias, i.e. general (relative) tendency to over- orunderpredict. Values are between -1 (complete under-prediction) and 1 (complete over-prediction) and 0 ideally. G: Absolute error of the median forecast relative to the Hub ensemble. H. Standard deviation of all WIS values for different horizons relative to the Hub ensemble.}\label{fig:agg-performance-ensemble-rel}
\end{figure}

@@

For deaths, contributions from the renewal model and crowd forecast together improved performance only for one week ahead predictions and showed an increasingly negative impact on performance for longer horizons (rel. WIS of the hub-ensemble-realised 1.01 two weeks ahead, 1.05 four weeks ahead, Tables \ref{tab:score-table-ensemble-2} and \ref{tab:score-table-ensemble-4}). Individual contributions from both the renewal model and the crowd forecast were largely negative, while a version of the Hub ensemble with only the convolution model included would have performed consistently better across all forecast horizons (with the positive impact increasing for longer horizons). This is especially interesting as the convolution model performed consistently worse than the pre-existing Hub ensemble (Figure \ref{fig:agg-performance-all}) and especially worse for longer horizons.

We also considered the impact of our contributions on a version of the Hub ensemble constructed by taking the quantile-wise mean, rather than the median.
General trends were similar, with the notable exception of the convolution model, which had a consistently positive impact on the median ensemble, but a mixed and mostly slightly negative impact on the mean ensemble (Figures \ref{fig:agg-performance-ensemble}B and \ref{fig:agg-performance-ensemble-mean}B). This may happen if a model is more correct directionally relative to the pre-existing ensemble, but overshoots in absolute terms, thereby moving the ensemble too far. For both the mean and the median ensemble, changes in performance from adding or removing models were of a similar order of magnitude, suggesting that at least in this instance, with a relatively small ensemble size, the median ensemble was not necessarily more `robust' to changes than the mean ensemble. However, the ensemble version with all our forecasts included (``hub-ensemble-with-all'') tended to perform relatively better for the median ensemble than the mean ensemble, suggesting that adding more models may be more beneficial or `safer' for the median than for the mean ensemble as directional errors can more easily cancel out than errors in absolute terms.

@@

\begin{figure}[H]
\includegraphics[width=1\linewidth,]{../analysis/plots/difference-wis-crowd-ensemble} \caption{Density plot with the difference in WIS between the Crowd forecast and the Hub ensemble (values below zero mean better performance of the Crowd forecasts).}\label{fig:distribution-scores-differences}
\end{figure}

@@

\begin{figure}[H]
\includegraphics[width=1\linewidth,]{../analysis/plots/aggregate-performance-rel-ensemble-mean-v4} \caption{Visualisation of aggregate performance metrics across forecast horizons for the different versions of the Hub mean ensemble. “Hub-ensemble” *excludes* all our models, Hub-ensemble-all *includes* all of our models, “Hub-ensemble-real” is the actual hub-ensemble observed in reality, which includes the renewal model and the crowd forecasts, but ont the convolution model. Values (except for Bias) are computed as differences to the Hub ensemble which excludes our contributions. For Coverage, this is an absolute difference, for other metrics this is a percentage difference. A, B: mean weighted interval score (WIS) across horizons relative to the Hub ensemble (lower values indicate better performance). C, D: Empirical coverage of the 50\% and 90\% prediction intervals minus empirical coverage observed for the Hub ensemble. E: Dispersion relative to the dispersion of the Hub ensemble. Higher values mean greater dispersion of the forecast and imply ceteris paribus a worse score. F: Bias, i.e. general (relative) tendency to over- orunderpredict. Values are between -1 (complete under-prediction) and 1 (complete over-prediction) and 0 ideally. G: Absolute error of the median forecast relative to the Hub ensemble. H. Standard deviation of all WIS values for different horizons relative to the Hub ensemble.}\label{fig:agg-performance-ensemble-mean}
\end{figure}

@@

In the original analysis, cases and deaths were scored on different periods, as the convolution model was only added later. This sensitivity shows performance of all models restricted to the period from October 14 2020 until March 1st 2021 where all models were available.

\begin{figure}[H]
\includegraphics[width=1\linewidth,]{../analysis/plots/aggregate-performance-all-late-period-v4} \caption{Visualisation of aggregate performance metrics across forecast horizons only for the period from October 14th 2020 on where all models were available. A, B: mean weighted interval score (WIS, lower indicates better performance) across horizons. WIS is decomposed into its components dispersion, over-prediction and under-prediction. C: Empirical coverage of the 50\% prediction intervals (50\% coverage is perfect). D: Empirical coverage of the 90\% prediction intervals. E: Dispersion (same as in panel A, B). Higher values mean greater dispersion of the forecast and imply ceteris paribus a worse score. F: Bias, i.e. general (relative) tendency to over- or underpredict. Values are between -1 (complete under-prediction) and 1 (complete over-prediction) and 0 ideally. G: Absolute error of the median forecast (lower is better). H. Standard deviation of all WIS values for different horizons}\label{fig:agg-performance-all-late}
\end{figure}

>>>>

Data on daily reported test positive cases and deaths linked to COVID-19 were provided by the organisers of the German and Polish Forecast hub. Until December 14th, 2020, these data were sourced from the European Centre for Disease Control (33). After ECDC stopped publishing daily data, observations were sourced from the Robert Koch Institute (RKI) and the Polish Ministry of Health for the remainder of the submission period (34). These data are subject to reporting artefacts, (such as for example delayed case reporting in Poland on the 24th November, (35)), changes in reporting over time, and variation in testing regimes (for example in Germany from the 11th of November on, (36)). The ECDC data as well as the data published by the Polish Ministry of Health were also subject to data revisions, although most of them (with a notable exception of a data update for October 12 2020 in Germany) only affected daily, not weekly data (see Figures \ref{fig:daily-truth-update} and \ref{fig:weekly-truth-update}).

@@

Our crowd forecasts were created as an ensemble of forecasts made by individual participants every week through a web application (\url{https://cmmid-lshtm.shinyapps.io/crowd-forecast/}). Weekly forecasts had to be submitted before Tuesday 12pm every week, but participants were asked to only use any information or data that was already available by Monday night. The application was built using the \texttt{shiny} and \texttt{golem} R packages (37,38) and is available in the \texttt{crowdforecastr} R package (31). To make a forecast in the application participants could select a predictive distribution (with the default being log-normal) to represent the probability that the forecasted quantity took certain values. Median and width of the uncertainty could be adjusted by either interacting with a figure showing their forecast or providing numerical values (see screenshot in Figure \ref{fig:screenshot} in the SI). The default shown was a repetition of the last known observation with constant uncertainty around it computed as the standard deviation of the last four changes in weekly log observed forecasts (i.e.~as \(\sigma(log(value4) - log(value3), log(value3) - log(value2), \ldots )\)). A comparison of the crowd forecasts against the default baseline shown in the application is displayed in Figure \ref{fig:compare-forecasters} in the Appendix.
Our interface also allowed participants to view past observations based on the hub data, as well as their forecasts, on a logarithmic scale and presented additional contextual COVID-19 data sourced from (39). These data included, for example, notifications of both test positive COVID-19 cases and COVID-19 linked deaths and the number of COVID-19 tests conducted over time. From November 26 2020 on we displayed weekly small reports with a visualisation of past forecasts and scores on our website, epiforecasts.io.

Forecasts were stored in a Google Sheet and downloaded, cleaned and processed every week for submission to the Forecast Hub. If a forecaster had submitted multiple predictions for a single target, only the latest submission was kept. Information on the chosen distribution as well as the parameters for median and width were used to obtain the required set of 23 quantiles from that distribution. Forecasts from all forecasters were then aggregated using an unweighted quantile-wise mean (i.e., the \(\alpha\)-quantile of the ensemble is given by the mean of all submitted \(\alpha\)-quantiles). To avoid issues with users trying out the app and submitting a random forecast, we required that a forecaster needed to make a forecast for at least two targets for a given forecast in order to be included in the crowd forecast ensemble. On a few occasions we deleted forecasts that were clearly the result of a user or software error (such as for example forecasts that were zero everywhere).

individual forecasts were assessed as clearly erroneous by visual inspection and subsequently removed before aggregation and were excluded from the submission as well as the analysis.

@@

All forecasting approaches, including the Hub ensemble, were overconfident, i.e.~they showed lower than nominal coverage (meaning that 50\% (90\%) prediction intervals generally covered less than 50\% (90\%) of the actually observed values) (Figure \ref{fig:agg-performance-all}C and \ref{fig:agg-performance-all}D). Coverage for all forecasts deteriorated with increasing forecast horizon, indicating that all forecasting approaches struggled to quantify uncertainty appropriately for case forecasts. This was especially an issue for crowd forecasts, which had markedly shorter prediction intervals (i.e., narrower and more confident predictive distributions) than other approaches (Figure \ref{fig:agg-performance-all}E) and only showed a small increase in uncertainty across forecast horizons. The crowd forecasts prediction intervals were also noticeably narrower than the default baseline shown to forecasters in the application (see Figure \ref{fig:compare-forecasters}).

In spite of good performance in terms of the absolute error (Figure \ref{fig:agg-performance-all}G), the narrow forecast intervals led to forecasts which were severely overconfident (covering only 36\% and 55\% of all observations with the 50\% and 90\% prediction intervals of all forecasts made at a two week forecast horizon, and only 5\% and 38\% four weeks ahead) (Figure \ref{fig:agg-performance-all}C,D and Tables \ref{tab:score-table-2} and \ref{tab:score-table-4}). Despite worse performance in terms of absolute error (Figure \ref{fig:agg-performance-all}G), the renewal model achieved better calibration (comparable to the Hub ensemble), as uncertainty increased rapidly across forecast horizons. The crowd forecasts, on the other hand, showed a smaller bias than the renewal model, but were overconfident.

@@

Generally, trends in overall performance were broadly similar across locations (Figures \ref{fig:performance-locations} and \ref{fig:performance-locations-rel}).
Due to the differing population sizes and numbers of notifications in Germany and Poland absolute scores were difficult to compare directly. However, relative to the Hub ensemble, the crowd forecasts performed noticeably better in Germany than in Poland and the renewal model better in Poland than in Germany (Figures \ref{fig:performance-locations-rel}A, \ref{fig:performance-locations-rel}G, \ref{fig:agg-performance-all-Germany}, \ref{fig:agg-performance-all-Poland}).

@@

\begin{figure}[H]
\includegraphics[width=1\linewidth,]{../analysis/plots/aggregate-performance-rel-ensemble-v4} \caption{Visualisation of relative aggregate performance metrics across forecast horizons for the different versions of the Hub median ensemble. “Hub-ensemble”   extit{excludes} all our models, Hub-ensemble-all    extit{includes} all of our models, “Hub-ensemble-realised” is the actual hub-ensemble observed in reality, which includes the renewal model and the crowd forecasts, but not the convolution model. A, B: mean weighted interval score (WIS) across horizons relative to the Hub ensemble (lower values indicate better performance). C, D: Empirical coverage of the 50\% and 90\% prediction intervals minus empirical coverage observed for the Hub ensemble. E: Dispersion relative to the dispersion of the Hub ensemble. Higher values mean greater dispersion of the forecast and imply ceteris paribus a worse score. F: Bias, i.e. general (relative) tendency to over- orunderpredict. Values are between -1 (complete under-prediction) and 1 (complete over-prediction) and 0 ideally. G: Absolute error of the median forecast relative to the Hub ensemble. H. Standard deviation of all WIS values for different horizons relative to the Hub ensemble.}\label{fig:agg-performance-ensemble-rel}
\end{figure}

@@

For deaths, contributions from the renewal model and crowd forecast together improved performance only for one week ahead predictions and showed an increasingly negative impact on performance for longer horizons (rel. WIS of the Hub-ensemble-realised 1.01 two weeks ahead, 1.05 four weeks ahead, Tables \ref{tab:score-table-ensemble-2} and \ref{tab:score-table-ensemble-4}). Individual contributions from both the renewal model and the crowd forecast were largely negative, while a version of the Hub ensemble with only the convolution model included would have performed consistently better across all forecast horizons (with the positive impact increasing for longer horizons). This is especially interesting as the convolution model performed consistently worse than the pre-existing Hub ensemble (Figure \ref{fig:agg-performance-all}) and especially worse for longer horizons.

We also considered the impact of our contributions on a version of the Hub ensemble constructed by taking the quantile-wise mean, rather than the median.
General trends were similar, with the notable exception of the convolution model, which had a consistently positive impact on the median ensemble, but a mixed and mostly slightly negative impact on the mean ensemble (Figures \ref{fig:agg-performance-ensemble-rel}B and \ref{fig:agg-performance-ensemble-mean}B). This may happen if a model is more correct directionally relative to the pre-existing ensemble, but overshoots in absolute terms, thereby moving the ensemble too far. For both the mean and the median ensemble, changes in performance from adding or removing models were of a similar order of magnitude, suggesting that at least in this instance, with a relatively small ensemble size, the median ensemble was not necessarily more `robust' to changes than the mean ensemble. However, the ensemble version with all our forecasts included (``hub-ensemble-with-all'') tended to perform relatively better for the median ensemble than the mean ensemble, suggesting that adding more models may be more beneficial or `safer' for the median than for the mean ensemble as directional errors can more easily cancel out than errors in absolute terms.

@@

\begin{figure}[H]
\includegraphics[width=1\linewidth,]{../analysis/plots/difference-wis-crowd-ensemble} \caption{Density plot with the difference in WIS between the Crowd forecast and the Hub ensemble (values below zero mean better performance of the Crowd forecasts) for a 2 week ahead forecast horizon.}\label{fig:distribution-scores-differences}
\end{figure}

\begin{figure}[H]
\includegraphics[width=1\linewidth,]{../analysis/plots/difference-wis-crowd-renewal} \caption{Density plot with the difference in WIS between the Crowd forecast and the Renewal model (values below zero mean better performance of the Crowd forecasts) for a 2 week ahead forecast horizon.}\label{fig:distribution-scores-differences-renewal}
\end{figure}

\newpage

@@

\begin{figure}[H]
\includegraphics[width=1\linewidth,]{../analysis/plots/aggregate-performance-rel-ensemble-mean-v4} \caption{Visualisation of aggregate performance metrics across forecast horizons for the different versions of the Hub mean ensemble. “Hub-ensemble”     extit{excludes} all our models, Hub-ensemble-all    extit{includes} all of our models, “Hub-ensemble-realised” is the actual hub-ensemble observed in reality, which includes the renewal model and the crowd forecasts, but ont the convolution model. Values (except for Bias) are computed as differences to the Hub ensemble which excludes our contributions. For Coverage, this is an absolute difference, for other metrics this is a percentage difference. A, B: mean weighted interval score (WIS) across horizons relative to the Hub ensemble (lower values indicate better performance). C, D: Empirical coverage of the 50\% and 90\% prediction intervals minus empirical coverage observed for the Hub ensemble. E: Dispersion relative to the dispersion of the Hub ensemble. Higher values mean greater dispersion of the forecast and imply ceteris paribus a worse score. F: Bias, i.e. general (relative) tendency to over- orunderpredict. Values are between -1 (complete under-prediction) and 1 (complete over-prediction) and 0 ideally. G: Absolute error of the median forecast relative to the Hub ensemble. H. Standard deviation of all WIS values for different horizons relative to the Hub ensemble.}\label{fig:agg-performance-ensemble-mean}
\end{figure}

@@

In the original analysis, cases and deaths were scored on different periods, as the convolution model was only added later. This sensitivity shows performance of all models restricted to the period from December 14 2020 until March 1st 2021 where all models were available.

\begin{figure}[H]
\includegraphics[width=1\linewidth,]{../analysis/plots/aggregate-performance-all-late-period-v4} \caption{Visualisation of aggregate performance metrics across forecast horizons only for the period from December 14th 2020 on where all models were available. A, B: mean weighted interval score (WIS, lower indicates better performance) across horizons. WIS is decomposed into its components dispersion, over-prediction and under-prediction. C: Empirical coverage of the 50\% prediction intervals (50\% coverage is perfect). D: Empirical coverage of the 90\% prediction intervals. E: Dispersion (same as in panel A, B). Higher values mean greater dispersion of the forecast and imply ceteris paribus a worse score. F: Bias, i.e. general (relative) tendency to over- or underpredict. Values are between -1 (complete under-prediction) and 1 (complete over-prediction) and 0 ideally. G: Absolute error of the median forecast (lower is better). H. Standard deviation of all WIS values for different horizons}\label{fig:agg-performance-all-late}
\end{figure}
